\chapter{Evaluation}
\label{ch:eval}

The previous chapter elaborates on a prototypical implementation of the concept
introduced in Chapter \ref{ch:concept}. Its goal is to enable fast serializable
database transactions through the use of novel byte-addressable non-volatile
memory. Based on the implemented prototype, this chapter attempts to determine
the effectiveness of its concept.

It has been shown, that MMDB can benefit from NVRAM \cite{bailey2013exploring,
schwalb2016hyrise, oukid2017data, andrei2017sap}. However, serializable
transactions are still traded in favor of performance, by default. If the
performance gain from NVRAM can be leveraged to compensate for serialization
overhead, then serializability could be adopted as the default isolation level.
Therefore, the aim of this chapter is to determine whether the approach of this
work is worth pursuing.

An important measure in this matter is transaction throughput, as it is known to
degrade due to higher abort rates in stronger isolation levels. For that purpose,
it is investigated whether transaction throughput with serializability can be as
high as in non-serializable systems and scale accordingly.

The evaluation is based on a performance comparison between the implemented
prototype, labeled as \emph{Midas}\footnote{The name is taken from Greek
mythology. According to the legend, King Midas was given the ability to turn
everything he touches into gold. This resembles the idea of making information
non-volatile as in NVRAM.}, and the NVRAM-based in-memory KVS \emph{Echo}
introduced in Chapter \ref{ch:kvs-nvram}. There are two benchmarks: a baseline
benchmark and a transaction throughput benchmark. The throughput benchmark is
used to compare the transaction throughput of both stores and how well it scales
with additional processors. The baseline benchmark captures average latencies of
single operations for each store, which are helpful to accomodate for
differences in the throughput benchmark.

At first, however, there a couple of challenges to discuss. Next, the system
configuration used for the evaluation is described. Subsequently, both baseline
and transaction throughput benchmarks are evaluated. The chapter concludes with
a brief summary.

\section{Challenges}
\label{ch:eval-challenges}
\input{contents/eval/challenges.tex}

\section{System Configuration}
\label{ch:eval-sysconfig}
\input{contents/eval/sysconfig.tex}

\section{Baseline}
\label{ch:eval-baseline}
\input{contents/eval/baseline.tex}

\section{Transaction Throughput}
\label{ch:eval-ttp}
\input{contents/eval/ttp.tex}

\section{Summary}
\label{ch:eval-summary}
\input{contents/eval/summary.tex}
