% This chapter is dedicated to determine whether MMDB can leverage NVRAM to make
% the isolation level of serializability affordable. For that purpose, the
% serializable KVS from Chapter \ref{ch:impl}, Midas, and the non-serializable KVS
% Echo are compared in terms of transaction throughput and scalability.

In 3 out of 4 scenarios, the original implementation of \midas performs
significantly worse than \echo. Especially in high-contention scenarios, it
shows much higher abort rates, lower throughput, and insignificant speedup.
However, at low contention, \midas consistently achieves higher throughput than
\echo. This is a promising result as it shows that, even with serializability,
MVCC can continue to excel in low-contention scenarios. However, despite
negligible abort rates, which are on par with Echo, \midas shows poor
scalability. This is due to the non-concurrent hash table implementation which
requires a global lock for synchronization.

Since the concrete index implementation is not part of the synchronization
concept, \midasopt is used to approximate the performance of a lock-free
concurrent hash table. \midasopt has significantly higher throughput and scales
well up until 16 cores, while maintaining comparable abort rates. However, this
result may be too optimistic because concurrent data structures are often slower
than their non-synchronized counterparts. Therefore, it is conceivable that
\midas with a concurrent index may perform worse than \midasopt but still better
than \echo.

\echo has much lower abort rates in all scenarios, but only provides snapshot
isolation. Moreover, it fails to leverage the advantage and hardly scales beyond
8 cores. Possible reasons for that include inappropriate synchronization and
inefficient NVRAM management.

In total, the benchmark shows that, in certain scenarios, serializability may
become affordable with NVRAM after all. Except in very high contention
scenarios, \midasopt can provide better throughput and scale better than \echo.
Moreover, if contention is low, even \midas can perform better than its
non-serializable counterpart \echo.
