\chapter{A Key-Value Store with Serializable Transactions for NVRAM}
\label{ch:concept}

The aim of this thesis is to design a \ac{KVS} with affordable serializable
transactions by exploiting the benefits of \ac{NVRAM}. In order to provide the
necessary groundwork, the previous chapters give an overview on recent research
in \ac{NVRAM}, \ac{KVS}, and concurrency control in databases. It is further
pointed out that, as of this writing, there appears to be no previous work on
leveraging \ac{NVRAM} for concurrency control in \ac{KVS} or \ac{MMDB},
respectively. While recent works primarily see \ac{NVRAM} as a means to reduce
recovery overhead, this thesis explores a different approach.

As mentioned in Chapter \ref{ch:kvs}, many transaction processing systems do not
support or encourage serializable transactions due to severe performance
degradations. Therefore, the idea is to use the benefits of \ac{NVRAM} to make
serializable transactions affordable. While this may not provide the highest
possible transaction throughput, the aim is to achieve performance on a par with
non-serializing solutions for traditional storage. In other words, instead of
increasing maximum performance, this work attempts to increase performance with
maximum consistency.

\ac{NVRAM} is especially significant for memory-resident databases as data no
longer need to be copied to a much slower storage device for recoverability.
Also, restarts can be near-instantaneous as all data are already in memory and
need not be fetched. Therefore, the proposed concept is exclusively targeted at
\ac{MMDB}. Given the vast complexity of fully-featured in-memory \ac{DBMS}, it
seems appropriate, for an initial study, to resort to much more manageable
\ac{KVS}. Based on whether the approach turns out to work well for \ac{KVS}, it
may still be applied on \ac{MMDB} in future work.

This chapter presents the concept for an \ac{NVRAM}-aware \ac{KVS} with
serializable transactions. After a brief overview in the next section,
follow-up sections outline the architecture, concurrency control, and
consistency measures.

\section{Overview}
\label{ch:concept-overview}
\input{contents/concept/overview}

\section{System Architecture}
\label{ch:concept-system}
\input{contents/concept/system}

\section{Key-Value Store Design}
\label{ch:concept-kvs}
\input{contents/concept/kvs}

\section{Crash Consistency}
\label{ch:concept-consistency}
\input{contents/concept/consistency}

\todo[inline]{Insert concluding example: e.g. structure w/ numbered arrows}
